---
title: "Characterizing Mysid (Crustacea: Mysida) swarms and their co-occurrence with gray whales (*Eschrichtius robustus*) in northwest Washington State"
author: "Liz Allyn"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Eastern North Pacific gray whales (*Eschrictius robustus*) migrate between their breeding grounds in Baja California, Mexico, and their feeding grounds in the Bering Sea every year. Their movement and distribution within that range during the feeding season is related to the quality of foraging opportunities and the distribution of prey species (Moore, Grebmeier & Davies, 2003; Bradford et al., 2012; Feyrer & Duffus, 2015; Stimmelmayr & Gulland, 2020; Hildebrand et al., 2022). A smaller group of whales called the Pacific Coast Feeding Group (PCFG) is known to use the area from northern California, USA to British Columbia, Canada more frequently during the feeding season instead of migrating all the way to the Bering Sea (Calambokidis et al., 2002).

PCFG whales exhibit a wide array of variability in foraging strategies and interannual site use (Darling, Keogh & Steeves, 1998; Scordino et al., 2017). Four discrete foraging strategies have been described in the PCFG range: benthic foraging on amphipods, epibenthic foraging on mysid swarms, pelagic foraging on crab larvae (primarily Porcellanidae), and shallow water foraging for ghost shrimp (Callianassa spp) (Nerini, 1984; Kvitek & Oliver, 1986; Duffus, 1996). This variability poses challenges for management, especially around defining the PCFG range, estimating the size of the group, determining sources of recruitment, and identifying critical habitat. As a result, many research groups have recently focused on identifying and characterizing the relationships between gray whales and their prey to increase our understanding of the PCFG and the role they play in the ecosystem.

In northwest Washington, gray whales exhibit spatial and temporal variability in their use of coastal areas during the feeding season, and the number of whales present in the region can also vary greatly within a season and between years (Scordino et al., 2017). Gray whale distribution has been related to prey availability in other research areas of the PCFG range. Much of this research has focused on the relationship between gray whales and mysid shrimp, which are a key prey item in Clayoquot Sound, BC, Canada and Port Orford, OR, USA. In Clayoquot Sound, researchers have documented close relationships between gray whale foraging and mysid abundance, including both a top-down effect of gray whale foraging on mysid abundance and an effect of mysid abundance on gray whale distribution in the area (Burnham, 2015; Burnham & Duffus, 2016, 2018). In Port Orford, studies have documented a threshold foraging response in gray whales where feeding behaviors were only initiated given a large enough abundance of high-calorie prey items (Hildebrand, Bernard & Torres, 2021; Hildebrand et al., 2022). It is likely that similar dynamics are at play in Washington, but no studies have yet documented the composition or abundance of mysid swarms in the area, or how they might influence gray whale behavior.

This study aimed to characterize the demographics of mysid swarms in northwest Washington and examine the co-occurrence of mysid swarms with feeding gray whales. We expected that gray whale presence would be predicted by the abundance of mysids sampled.

## Methods

### Study Area

Small boat surveys were conducted in northwest Washington State as described in Scordino et al. (2017). The primary purpose of these surveys was to document gray whale sightings and collect photographs for individual photo identification. During the gray whale feeding season in 2019 and 2020, plankton tows were conducted concurrently at discrete sites along the established survey route to collect preliminary information about the prey base available to gray whales and the demographics of mysid swarms in this region (Figure 1 - not included in figure limit for this report).

### Sample Collection

Sites along the established gray whale survey route were chosen to conduct sampling for mysid shrimp presence and demographic composition. Seven sites in the Strait of Juan de Fuca and seven sites in the Pacific Ocean were surveyed. All sample sites were close to shore in depths of less than 50 ft and were selected based on areas where gray whales have previously been documented to feed (Scordino et al., 2017) and a desire to stratify sampling sites throughout the study region. Exact sample locations moved closer or further from the shore throughout the season to consistently sample along the edge of the expanding and contracting kelp line, as navigation within kelp beds was not possible with our research vessel, and the habitat immediately surrounding kelp beds in this region is often not suitable for mysids. The southernmost site sampled in 2019 was the South of Bodelteh Islands site, but sampling effort was moved to the Ozette Island site in 2020 to expand the sampling area further south. We aimed to collect one sample per site each month, but sampling effort was highly variable due to a number of factors, including lost or malfunctioning equipment, inclement weather, or the need to prioritize other primary research objectives. Sampling occurred during the gray whale feeding season from June to November in 2019 and from June to September in 2020.

Samples were collected using a Bongo-style net manufactured by Sea-Gear Corporation (Melbourne, FL, USA) similar to ones used by other regional research groups (Burnham, 2015; Hildebrand et al., 2022). Twin hoops measured 30 cm in diameter with a 3:1 mouth to length ratio and a 500-µm mesh. At each site, the net was lowered over the side by hand until it reached the bottom (confirmed through the collection of benthic species, substrate sediments, estimated line out from depth sounder, and the ability of the technician to feel the bottom), then was towed slowly (approximately 1 kt through the water, accounting for current speed and direction) near the bottom for 30 seconds and retrieved as fast as could be achieved by hand (approximately 0.6 m s-1). This procedure was repeated for two tows at each sample site, and contents of the cod end from each tow were poured into a single 500-mL container. If the sample did not contain mysids or other known gray whale prey items then the sampling effort was recorded but the sample was not retained. Samples were stored in 70% ethanol before processing.

### Sample Processing

For each sample, species identifications of all mysids were conducted according to Kathman (1986) using a 40x dissecting microscope. Mysids were measured from the tip of the rostrum to the tip of the telson and the species, sex, maturity, and gravidity were identified. Mysids were designated as "Unknown" species if they couldn't be identified due to a damaged or absent telson. Mysids were assigned a sex based on the presence of a brood sac for females or a reproductive appendage for males, and mysids that did not have either characteristic were designated as (insert new designation here). Seven samples with high mysid density (visually estimated as containing \>1000 mysids) were approximated by visually estimating the total count of mysids in the sample, collecting a subsample of 100 mysids from that sample, and identifying the species, sex, length, maturity, and gravidity of the subsample. The demographic composition of the subsample was then extrapolated to characterize the full sample. These tows that were characterized via subsampling were included in all subsequent analyses and can be inspected separately in our publicly available data repository (see Allyn et al., 2023). Other non-mysid organisms collected during sampling were identified to the lowest taxonomic level possible but were not included in subsequent analyses. These data are available in the data repository or upon request.

### Gray Whale Surveys

Gray whale sightings were recorded throughout the study period during small boat surveys of northwest Washington using the methods described in detail in Scordino et al. (2017). In summary, each gray whale sighted during surveys was photographed using digital SLR cameras with a 100 -- 400 mm lens and the time and location of the sighting was recorded. Photographs of gray whales were sent to Cascadia Research Collective (Olympia, WA, USA) for inclusion in their catalog of individually identified gray whales from the west coast of the contiguous USA and Canada. Each whale photographed was processed according to establish procedures and was either matched to an existing whale and identification number in the catalog, assigned a new identification number if no match could be made, or left unidentified if photos of the whale were of insufficient quality (Calambokidis et al., 2002; Calambokidis, Pérez & Laake, 2019). Individual identifications were used to determine the total number of individual whales observed in the survey area, since whales can be sighted more than once during survey effort during the same day or on subsequent days throughout the season. Surveys were conducted under Marine Mammal Protection Act research permits granted by NMFS to author JJS (23970 and 19430).

### Analysis

Data analysis and visualization was conducted using Microsoft Excel and R version 4.1.1 (R Core Team, 2021). All prey sample data underlying the results presented in the study are available from Mendeley Data (Allyn et al., 2023). Whale sighting data is available upon request from the Makah Fisheries Management Department of the Makah Tribe (see Data Declaration for contact information).

Mysid swarm size, species, sex, and maturity compositions were summarized and described. A generalized linear mixed hurdle model was constructed to evaluate if the presence of whales near sample sites during sampling could be predicted by the abundance of mysids sampled at that location as a proxy for prey availability in the area.

## Results

```{r message = F}
# source("~Users/lizallyn/Documents/GitHub/Project-Mysids/Project Mysids.Rmd") # not working

library(ggplot2)

# Read in Data
data <- read.csv("Er prey analysis for R fixed whale presence.csv")

data$Site <- factor(data$Site, levels = c("Chito Beach", "Bullman Beach", "Seal and Sail", "Sail River", "First Beach", "Koitlah", "Slant Rock", "Skagway Rocks", "Anderson Rocks", "Portage Head", "Duk Point", "North Bodelteh", "South Bodelteh", "Ozette Island"))

## Inspect data distribution

hist(data$whalecount) # zero-inflated
# will need hurdle model to account for this

## Check for overdispersion in nonzero counts

# pull out non-zero counts
nonzero <- data[which(data$whalecount > 0),]

library(VGAM)

# construct zero-truncated Poisson model with mysid count
m1 <- vglm(whalecount ~ MysidCount, data = nonzero, family = "pospoisson")
summary(m1)

# check for overdispersion
pchisq(sum(residuals(m1,type = "pearson")^2),nrow(nonzero)-2,lower.tail = FALSE)
# p = 0.000005, overdispersed

# fit with ZT neg binom instead
m2 <- vglm(whalecount ~ MysidCount, data = nonzero, family = posnegbinomial)
summary(m2)

## Look for weird things in data
# hat values
library(faraway)
halfnorm(hatvalues(m2), labs = nonzero$Sample)
# shows points with very high mysid counts - makes sense
# model structure accounts for a lot of other weird things, so I don't
# think we need any other diagnostics yet

## Construct hurdle model with ZT neg binom and assess Site as random effect.
# global model at this point includes Month as well

library(glmmTMB)
library(TMB)
library(bbmle) # AICtab

glmm.1 <- glmmTMB(data = data, whalecount ~ MysidCount + StudyMon + (1|Site), family = truncated_nbinom1, ziformula = ~.)
summary(glmm.1)
glmm.0 <- glmmTMB(data = data, whalecount ~ MysidCount + StudyMon, family = truncated_nbinom1, ziformula = ~.)
summary(glmm.0)
AICtab(glmm.1, glmm.0)
# delta AIC says no random effect better by 1, not large enough to exclude
# Site makes sense according to structure of data.

## Assess fixed effect structure

library(knitr)
library(plyr)
library(dplyr)

data$Site <- as.factor(data$Site)
glmm.mm <- glmmTMB(data = data, whalecount ~ MysidCount + StudyMon + (1|Site), family = truncated_nbinom1, ziformula = ~.)
summary(glmm.mm)
glmm.m <- glmmTMB(data = data, whalecount ~ MysidCount + (1|Site), family = truncated_nbinom1, ziformula = ~.)
summary(glmm.m)
# null model not currently running? error below.
glmm.0 <- glmmTMB(data = data, whalecount ~ 1 + (1|Site), family = truncated_nbinom1, ziformula = ~.)
# Error in eigen(h) : infinite or missing values in 'x'
# summary(glmm.0)

kable(AICtab(glmm.mm, glmm.m))
# including month is worse (deltaAIC = 2.4) - not a lot worse, but I'm going to exclude anyway. In another scenario I might choose to do weighted predictions with both models.

# remember to do the AIc comparison when all three models are running!!

## Predictions

plot(data$MysidCount, predict(glmm.m, type = "response"))
# looks good now! type = "response" undoes the link function transformation! Key!!

# model of the mean predictions by hand bc model not running above
null.res <- data %>%
  group_by(Site) %>%
  summarise(site.mean = mean(whalecount))

test <- join(data, null.res, by = 'Site')

theme.Speciesym <- theme_classic() +
  theme(plot.margin = margin(t=10,r=10,b=10,l=10),
        axis.title = element_blank(), 
        axis.title.x = element_text(color = "black", hjust = 0.5, vjust = 0, size = 12), 
        axis.title.y = element_text(hjust = 0.45, vjust = 2, color = "black", size = 12), 
        plot.title = element_blank(), 
        axis.text.x = element_text(size = 10, colour = "black"), 
        axis.text.y = element_text(size = 10, colour = "black"),
        axis.ticks.x = element_blank(), 
        legend.title = element_text(size = 12, colour = "black"), 
        legend.text = element_text(size = 10, colour = "black", face = "italic"), 
        legend.key.size = unit(1, "line")) 

scatterplot <- ggplot() + 
  geom_point(data = data, aes(x = MysidCount, y = whalecount, color = Site), size = 3, alpha = 0.5) + 
  geom_line(data = data, aes(x = MysidCount, y = predict(glmm.m, type = "response"), group = Site, color = Site)) +
  labs(x = "Mysid Count per Sample", y = "Number of Whales near Site during Sample") +
  theme.Speciesym
scatterplot

```

(Descriptive mysid composition results excluded for this report.)

### Model Selection

The whale count data was zero-inflated, so a hurdle model was constructed using glmmTMB (Brooks et al., 2017) with a binomial hurdle and zero-truncated negative binomial for the overdispersed non-zero count data. The equation for this model is

$$
f(y|p,\lambda) = 
\begin{cases}
   Pr(y=0) & (1-p)\\
   Pr(y>0) & p~*~\frac{\Gamma(y+n)p^n(1-p)^y}{\Gamma(n)y!(1-p^n)}\\
\end{cases}
$$ where;

-   *p* is the probability of observing a whale at a given sample site during sampling

-   *y* is the predicted count of whales present

-   *n* is the number of samples

A model was constructed with whale count as the response and mysid count as the predictor, with sample Site as a random effect to account for repeated measures. The model with the random effect of Site had a delta AIC of 1 compared to the model with no random effect, but including it makes sense for the structure of the data. This model was compared against the null intercept model using AIC.

We checked for goodness of fit with the mysid model using the DHARMa package (Hartig, 2022) as well as visual checks on the plotted model predictions.

(insert scatterplot with model predictions and mean for each site. Maybe also AIC table?)

## Discussion

The null model performed better than the model with mysid count as a predictor. This means that mysid counts in a nearby sample do not give us any more ability to predict the number of whales near the sample site than the average whale count for that site independent of the sample composition. This is somewhat surprising because we know that whales respond to the distribution of their prey in other nearby systems (Clayoquot Sound, BC and Port Orford, OR), so we expected to see a similar response here.

It's possible that our sampling procedure is not well designed to capture patterns at such a small spatial and temporal scale (the area around the sample site during the sampling activity). PCFG gray whales utilize a broad spectrum of prey species and foraging strategies, so their distribution during the feeding season may be indicative of overall prey availability and system productivity in a given region. Within that larger pattern of system productivity that the whales may be responding to, our mysid samples may not be able to capture a comprehensive snapshot of the prey available to whales.

The disconnect between whale distribution and mysid counts at this small scale could also be due to the patchiness of mysid swarms and the discrete nature of our sampling methods. Mysid swarms are veryy patchy, so our discrete plankton net sampling method may be missing crucial blooms of prey availability, or conversely sampling from a swarm that is more abundant than the surrounding area as a whole. In this way our sampling could easily be disconnected from the cues whales need to determine their foraging effort.

To address the patchy nature of the data collected in this study, I think it would be interesting to look at the data aggregated on a slightly larger spatial and temporal scale. From our whale surveys we are able to determine the total number of uniquely identified whales in the study region that we encountered on a given survey day. We could similarly summarize the average number of mysids caught per tow for each survey day, and see if there is any relatedness between these daily measures. In this analysis, the average daily mysids per tow would act as an indicator of the overall productivity of the area. We would then expect the number of whales in the area to increase as the average mysids increased, and decrease when there are few mysids in the area.

## References

Bradford AL, Weller DW, Punt
AE, Ivashchenko Y v., Burdin AM, van Blaricom GR, Brownell RL. 2012. Leaner
leviathans: Body condition variation in a critically endangered whale
population. *Journal of Mammalogy* 93:251--266. DOI:
10.1644/11-MAMM-A-091.1.

Brooks Mollie E., Kasper Kristensen, Koen J. van Benthem, Arni Magnusson, Casper W. Berg, Anders Nielsen, Hans J. Skaug, Martin Maechler and Benjamin M. Bolker (2017). glmmTMB Balances Speed and Flexibility Among Packages for
 Zero-inflated Generalized Linear Mixed Modeling. The R Journal, 9(2), 378-400. doi: 10.32614/RJ-2017-066.

Burnham RE. 2015.
Reproductive strategies conferring species dominance in marine mysid
(Peracarida, Mysida) species in coastal waters off Vancouver Island, BC. *Crustaceana*
88:1421--1438. DOI: 10.1163/15685403-00003502.

Burnham RE, Duffus DA. 2016.
Gray whale (Eschrichtius robustus) predation and the demise of amphipod prey
reserves in Clayoquot Sound, British Columbia. *Aquatic Mammals*
42:123--126. DOI: 10.1578/AM.42.2.2016.123.

Burnham R, Duffus D. 2018.
Patterns of predator-prey dynamics between gray whales (Eschrichtius robustus)
and mysid species in Clayoquot Sound. *Journal of Cetacean Research and
Management* 19:95--103.

Calambokidis J, Darling JD,
Deecke V, Gearin P, Gosho M, Megill W, Tombach CM, Goley D, Toropova C,
Gisborne B. 2002. Abundance, range and movements of a feeding aggregation of
gray whales (Eschrichtius robustus) from California to southeastern Alaska in
1998. *Journal of Cetacean Research and Management* 4:267--276.

Darling JD, Keogh KE, Steeves
TE. 1998. Gray whale (Eschrichtius robustus) habitat utilization and prey
species off Vancouver Island, BC. *Marine Mammal Science* 14:692--720. DOI:
10.1111/j.1748-7692.1998.tb00757.x.

Duffus DA. 1996. The
recreational use of grey whales in southern Clayoquot Sound, Canada. *Applied
Geography* 16:179--190. DOI: 10.1016/0143-6228(96)00002-1.

Feyrer LJ, Duffus DA. 2015.
Threshold foraging by gray whales in response to fine scale variations in mysid
density. *Marine Mammal Science* 31:560--578. DOI: 10.1111/mms.12178.

Hartig F (2022). \_DHARMa: Residual Diagnostics for Hierarchical (Multi-Level / Mixed) Regression Models\_.R package version 0.4.6, <https://CRAN.R-project.org/package=DHARMa>.

Hildebrand L, Bernard KS,
Torres LG. 2021. Do Gray Whales Count Calories? Comparing Energetic Values of
Gray Whale Prey Across Two Different Feeding Grounds in the Eastern North
Pacific. *Frontiers in Marine Science* 0:1008. DOI:
10.3389/FMARS.2021.683634.

Hildebrand L, Sullivan FA,
Orben RA, Derville S, Torres LG. 2022. Trade-offs in prey quantity and quality
in gray whale foraging. *Marine Ecology Progress Series* 695:189--202. DOI:
10.3354/meps14115.Kvitek RG, Oliver JS. 1986.
Side-scan sonar estimates of the utilization of gray whale feeding grounds
along Vancouver Island, Canada. *Continental Shelf Research* 6:639--654.
DOI: 10.1016/0278-4343(86)90028-2.

Moore SE, Grebmeier JM,
Davies JR. 2003. Gray whale distribution relative to forage habitat in the
northern Bering Sea: current conditions and retrospective summary. *Canadian
Journal of Zoology* 81:734--742. DOI: 10.1139/z03-043.

Nerini M. 1984. A Review of
Gray Whale Feeding Ecology. In: Jones M lou, Swartz SL, Leatherwood S eds. *The
Gray Whale: Eschrichtius Robustus*. Academic Press, 423--450.

Scordino JJ, Gosho M, Gearin
PJ, Akmajian A, Calambokidis J, Wright N. 2017. Individual gray whale use of
coastal waters off northwest Washington during the feeding season 1984--2011:
Implications for management. *Journal of Cetacean Research and Management*
16:57--69.

Stimmelmayr R, Gulland FMD.
2020. Gray Whale (Eschrichtius robustus) Health and Disease: Review and Future
Directions. *Frontiers in Marine Science* 7:963. DOI:
10.3389/FMARS.2020.588820/BIBTEX.
