---
title: "Project Mysids"
author: "Liz Allyn"
date: "`r Sys.Date()`"
output: word_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r message = F, echo = F}
```

# Characterizing Mysid (Crustacea: Mysida) swarms and their co-occurrence with gray whales (Eschrichtius robustus) in northwest Washington State

## Research Questions

### Descriptive:

-   What is the species composition of mysid shrimp in the study area?
-   What is the size, age, maturity composition of mysid shrimp in the
    study area?
-   How does mysid shrimp distribution vary by sample site?

### Modeling:

-   Does mysid shrimp density predict gray whale presence?

## Methods

### Data Organization

```{r}
library(ggplot2)

# Read in Data
data <- read.csv("Er prey analysis for R fixed whale presence.csv")

# Visualize the relationship
plot(data$MysidCount, data$whalecount)

data$Site <- factor(data$Site, levels = c("Chito Beach", "Bullman Beach", "Seal and Sail", "Sail River", "First Beach", "Koitlah", "Slant Rock", "Skagway Rocks", "Anderson Rocks", "Portage Head", "Duk Point", "North Bodelteh", "South Bodelteh", "Ozette Island"))

theme.Speciesym <- theme_classic() +
  theme(plot.margin = margin(t=10,r=10,b=10,l=10),
        axis.title = element_blank(), 
        axis.title.x = element_text(color = "black", hjust = 0.5, vjust = 0, size = 15), 
        axis.title.y = element_text(hjust = 0.45, vjust = 2, color = "black", size = 15), 
        plot.title = element_blank(), 
        axis.text.x = element_text(size = 13, colour = "black"), 
        axis.text.y = element_text(size = 13, colour = "black"),
        axis.ticks.x = element_blank(), 
        legend.title = element_text(size = 15, colour = "black"), 
        legend.text = element_text(size = 13, colour = "black", face = "italic"), 
        legend.key.size = unit(2, "line")) 

scatterplot <- ggplot(data = data, aes(x = MysidCount, y = whalecount)) + 
  geom_point(aes(color = Site), size = 3) + 
  labs(x = "Mysid Count per Sample", y = "Number of Whales near Site during Sample") +
  theme.Speciesym
scatterplot
# ggsave(plot = scatterplot, "scatterplot of whales by sample color by site.pdf", 
#        width = 10, height = 8, device = "pdf", dpi = 700)
```

This makes it look like there's no good reason to be looking for a relationship here...

### Model Construction

The response variable for this model is whale counts near sampling sites during sampling. Here we look at the count distribution with a histogram.

```{r}
hist(data$whalecount)
```

Since the data are counts, we make no assumption of normality from the start. The data are clearly zero-inflated from looking at the histogram, so we will need to use a Hurdle model.

Pull out non-zero counts and check for overdispersion.

```{r}
# pull out non-zero counts
nonzero <- data[which(data$whalecount > 0),]

# look at them
hist(nonzero$whalecount)
mean(nonzero$whalecount)
```

The data seem likely to be overdispersed - mostly 1, with just a few observations of up to 8 whales (which is likely to be outside CIs because the mean is 1.9). Below I check for overdispersion using deviance and Pearson's chi square test.

```{r message=F}
library(VGAM)
library(faraway)

# construct zero-truncated Poisson model with mysid count
m1 <- vglm(whalecount ~ MysidCount, data = nonzero, family = "pospoisson")
summary(m1)

# check for overdispersion
pchisq(sum(residuals(m1,type = "pearson")^2),nrow(nonzero)-2,lower.tail = FALSE)
# p = 0.000005, overdispersed

# fit with ZT neg binom instead
m2 <- vglm(whalecount ~ MysidCount, data = nonzero, family = posnegbinomial)
summary(m2)
```

Since the non-zero count data is overdispersed, we need to use the negative binomial distribution instead of Poisson in our hurdle model.

This is a Poisson hurdle - replace with appropriate ZTNegBinom!!
$$
f(y|p,\lambda) = 
\begin{cases}
   Pr(y=0) & (1-p)\\
   Pr(y>0) & p~*~\frac{e^{-\lambda}\lambda^y}{y!(1-e^{-\lambda})}
\end{cases}
$$

Next we will check for weird points and other diagnostics. We've settled on the model framework already because of the nature of the data, but good to check for weird things anyway.

```{r}
# plot residuals
plot(fitted(m2), residuals(m2)[,2])

# hat values
halfnorm(hatvalues(m2), labs = nonzero$Sample)
```
Residuals look very weird, I think that's ok though?
Hat values pulled out 20191106-4 which had 1400 mysids and 20200716-2 which had 1500 mysids, so those are pretty elevated counts.
This all seems fine, we're already using a model structure that accounts for a lot of these potential issues. 

Next we will construct the hurdle model manually as a two-part process.

```{r}
library(lme4)

# hurdle
m.0 <- glmer(whalepresence ~ MysidCount + (1|Site), data = data, family = binomial)
summary(m.0)

# non-zero count data
m.non0 <- glmer.nb(whalecount ~ MysidCount + (1|Site), data = nonzero)
summary(m.non0)
```

Here we use glmmTMB to construct the hurdle model with binomial zeros and zero-truncated negative binomial nonzeros, with random effects. Then we evaluate the random effect structure. Site is the only candidate effect. 

```{r}
library(glmmTMB)
library(TMB)
library(bbmle) # AICtab

glmm.1 <- glmmTMB(data = data, whalecount ~ MysidCount + scale(Month) + (1|Site), family = truncated_nbinom1, ziformula = ~.)
summary(glmm.1)
glmm.0 <- glmmTMB(data = data, whalecount ~ MysidCount + scale(Month), family = truncated_nbinom1, ziformula = ~.)
summary(glmm.0)
AICtab(glmm.1, glmm.0)
```

The model with site is not better than the model with no random effects, but it's not more than 2 dAIC worse, and makes sense for the data structure, so I'm deciding to keep it for now.



Fixed Effects: 
+ Mysid Count 
+ NR Count
+ HS Count
+ Mysid Avg Size 
+ Mysid Gravid Count
+ Month?
+ Temp?
+ Kelp Cover?

Look for effect of MysidCount first, then can look more detailed if that effect exists. 

## Fixed Effects

```{r}
library(knitr)
library(plyr)
data$Site <- as.factor(data$Site)
glmm.m <- glmmTMB(data = data, whalecount ~ scale(MysidCount) + (1|Site), family = truncated_nbinom1, ziformula = ~.)
summary(glmm.m)
glmm.0 <- glmmTMB(data = data, whalecount ~ 1 + (1|Site), family = truncated_nbinom1, ziformula = ~.)
# Error in eigen(h) : infinite or missing values in 'x'
summary(glmm.0)
kable(AICtab(glmm.0, glmm.m))

# model of the mean by hand bc not running
null.res <- data %>%
  group_by(Site) %>%
  summarise(site.mean = mean(whalecount))

# I could do this in 3 sec in Excel w/vlookup - assign site means to rows in data
data$site.mean <- null.res[which(null.res$Site == data$Site)]
null.res$Site %in% data$Site

test <- join(data, null.res, by = 'Site')

plot(data$MysidCount, predict(glmm.m))
```
The model of the mean is better than the one with MysidCount as a predictor...not sure where to go from here?
Tried to add month but inconsistent coverage of sites x months, so different lengths of data I think? Doesn't run at the moment. 

The predictions from these models are very unrealistic - lots of negative whales at high mysid density. Makes sense why this is such a bad model. I think this makes sense that it fit as a negative beta though from looking at the scatterplot - the trend really does seem negative. So I'm guessing it's a good fit to the data, but the data is a bad representation of the system dynamics. Though the plot below does not confirm this theory...

```{r}
theme.Speciesym <- theme_classic() +
  theme(plot.margin = margin(t=10,r=10,b=10,l=10),
        axis.title = element_blank(), 
        axis.title.x = element_text(color = "black", hjust = 0.5, vjust = 0, size = 12), 
        axis.title.y = element_text(hjust = 0.45, vjust = 2, color = "black", size = 12), 
        plot.title = element_blank(), 
        axis.text.x = element_text(size = 10, colour = "black"), 
        axis.text.y = element_text(size = 10, colour = "black"),
        axis.ticks.x = element_blank(), 
        legend.title = element_text(size = 12, colour = "black"), 
        legend.text = element_text(size = 10, colour = "black", face = "italic"), 
        legend.key.size = unit(1, "line")) 

scatterplot <- ggplot() + 
  geom_point(data = data, aes(x = MysidCount, y = whalecount, color = Site), size = 3, alpha = 0.5) + 
  geom_line(data = test, aes(x =  MysidCount, y = site.mean, color = Site)) +
  labs(x = "Mysid Count per Sample", y = "Number of Whales near Site during Sample") +
  theme.Speciesym
scatterplot
```



### Break - beyond this project below

Any kind of summary would be kind of like adding Day or Month as a random effect? Maybe first look at whales per day in total area with avg mysids per tow per day.

```{r}
# summarize by day to remove duplicate data

# pipe avg mysids per sample per day
summ.1 <- data %>%
  group_by(Date) %>%
  summarise(avg.mysids <- mean(MysidCount),
            unique.w <- mean(uniquewhal))

daily <- data.frame(summ.1)
colnames(daily) <- c("Date", "AvgMysids", "UniqueW")

# look at data
hist(daily$UniqueW)
plot(daily$AvgMysids, daily$UniqueW)

theme.Speciesym <- theme_classic() +
  theme(plot.margin = margin(t=10,r=10,b=10,l=10),
        axis.title = element_blank(), 
        axis.title.x = element_text(color = "black", hjust = 0.5, vjust = 0, size = 15), 
        axis.title.y = element_text(hjust = 0.45, vjust = 2, color = "black", size = 15), 
        plot.title = element_blank(), 
        axis.text.x = element_text(size = 13, colour = "black"), 
        axis.text.y = element_text(size = 13, colour = "black"),
        axis.ticks.x = element_blank(), 
        legend.title = element_text(size = 15, colour = "black"), 
        legend.text = element_text(size = 13, colour = "black", face = "italic"), 
        legend.key.size = unit(2, "line")) 

scatterplot.daily <- ggplot(data = daily, aes(x = AvgMysids, y = UniqueW)) + 
  geom_point(size = 3) + 
  labs(x = "Average mysid count per sample", y = "Unique whales in the study area") +
  theme.Speciesym
scatterplot.daily
ggsave(plot = scatterplot.daily, "scatterplot of unique whales by avg mysids.pdf",
       width = 10, height = 8, device = "pdf", dpi = 700)
```
Still zero-inflated. Looks like more of a trend too - almost never 0 whales when nonzero mysids around maybe?
Check for overdispersion

```{r}
# pull out nonzero data
daily.n0 <- daily[which(daily$UniqueW > 0),]

# construct zero-truncated Poisson model with mysid count
m5 <- vglm(UniqueW ~ AvgMysids, data = daily.n0, family = "pospoisson")
summary(m5)

# check for overdispersion
pchisq(sum(residuals(m5,type = "pearson")^2),nrow(daily.n0)-2,lower.tail = FALSE)
# overdispersed

# try the neg binom
m6 <- vglm(UniqueW ~ AvgMysids, data = daily.n0, family = "posnegbinomial")
summary(m6)

```

```{r}
# build a binom to frankenstein the thing bc the hurdle glmmTMB below isn't working

# create presence column
daily$Presence <- rep(NA, nrow(daily))
daily$Presence[which(daily$UniqueW == 0)] <- 0
daily$Presence[which(daily$UniqueW > 0)] <- 1


```


Data are overdispersed, do a neagtive binomial hurdle, but no random effects this time.

```{r}
library(nloptr)
# control parameters? from Whooping crane file
cntrl <- glmmTMBControl(optimizer = bobyqa, optCtrl = list(maxeval = 1e5))

glmm.daily.1 <- glmmTMB(data = daily, UniqueW ~ 1, family = truncated_nbinom1, ziformula = ~ AvgMysids, control = cntrl)
# Error in is.nloptr(ret) : length(lb) != length(x0)

summary(glmm.daily.1)
```


What if we look at region-wide per month. Average number of mysids per sample per day. Total number of unique whales in the study area per day that samples were taken. 

```{r}
library(dplyr)
whales <- read.csv("whales per day for R.csv")

# pipe avg mysids per sample per day
summ.1 <- data %>%
  group_by(Date) %>%
  summarise(avg.mysids <- mean(MysidCount))

# pull out whale days when samples were taken
whale.days <- subset(x = whales, subset = Date %in% data$Date)

# create data frame
summ.1$unique.w <- whale.days$Unique
summ.1$total.w <- whale.days$Whales

daily <- data.frame(summ.1)
colnames(daily) <- c("Date", "AvgMysids", "UniqueW", "TotalW")
```


