---
title: "Project Mysids"
author: "Liz Allyn"
date: "`r Sys.Date()`"
output: word_document
---

```{r message = F, echo = F}
library(knitr)
library(glmm)
library(lme4)
```

# Mysid distribution, demographics, and co-occurrence with gray whales on the NW coast of Washington State

## Research Questions

### Descriptive:
+ What is the species composition of mysid shrimp in the study area?
+ What is the size, age, maturity composition of mysid shrimp in the study area?
+ How does mysid shrimp distribution vary by sample site?

### Modeling:
+ Does mysid shrimp density predict gray whale presence?

## Methods


## Data Organization

```{r}
# Read in Data
data <- read.csv("Er prey analysis for R fixed whale presence.csv")

#


```


## Model Construction

Options:

Response:
+ Whale Presence or Absence (Bernoulli?)

Fixed Effects
+ Mysid Count
+ Mysid Avg Size
+ Mysid Gravid Prop
+ Mysid NR:HS Ratio (expect interaction w/count)
+ Month
+ Temp

Random Effects
+ Site
+ Date?

```{r}
# model creation

# intercept only
m0 <- glm(data = data, whale.presence ~ MysidCount, family = "poisson")
```

## Diagnostics

```{r}

```


